Probabilities and propensities: conversion experiences in online grocery shopping

### 2017-07-18 log.rmd
restructuring last section

## Mon 17 Jul 2017 11:46:32 BST
- still not sure how radically to restructure this. 
- the simplest order I can think of
    - A. Introduction - includes a,b,c from above
    - B. Patel section + opening description
    - C. Shopping and lists -- Miller, and list section
    - D. Archaeology of recommendations -- from 1984 -2007
    - E. Experimenting with probabilistic conversions
    - F. Conclusion
- will see if this structure works with a rough cut and paste
- re-ordering has gone ok to the end of D. But section E is really tricky -- this is the conversion event; do I include the machine learning stuff as part of D; or use it to be begin E?
    - E: shift from apriori to logistic regression; then think of good features; experimentalising/trials; infrastructural operationalisation; but then what about substitution; complementarity? All of this section is meant to be about probabilisation? And the propensities idea;    
- TODO: write short paragraph describing my overall social theory of lists (approx 200 words)
- new title: Probabilities in the data analytic conversion of social order
- TODO: play with the structuring of Section E. 
-

>>>>

## Mon 17 Jul 2017 11:46:32 BST
- still not sure how radically to restructure this. 
- the simplest order I can think of
    - A. Introduction - includes a,b,c from above
    - B. Patel section + opening description
    - C. Shopping and lists -- Miller, and list section
    - D. Archaeology of recommendations -- from 1984 -2007
    - E. Experimenting with probabilistic conversions
    - F. Conclusion
- will see if this structure works with a rough cut and paste
- re-ordering has gone ok to the end of D. But section E is really tricky -- this is the conversion event; do I include the machine learning stuff as part of D; or use it to be begin E?
    - E: shift from apriori to logistic regression; then think of good features; experimentalising/trials; infrastructural operationalisation; but then what about substitution; complementarity? All of this section is meant to be about probabilisation? And the propensities idea;    
- TODO: write short paragraph describing my overall social theory of lists (approx 200 words)
- new title: Probabilities in the data analytic conversion of social order
- TODO: play with the structuring of Section E.

## Tue 18 Jul 2017 07:18:46 BST
- adding Marcus on prototyping method into intro; 
- starting on section E: it will be rough
- how do I actually use the social theory of lists - what happens to the mapping? to the discrepancies between normative and actual, and to the siting of the list?
- TODO: write about this in the conclusion
- remapped last 12 pages a bit, taking into account the Popper argument and critiquing the Hansen a bit more;  see page 21 of July printout for summary; still feels like need more experiments or prototypes; particularly for the model itself; 
- TODO: rewrite section on logistic regression in terms of the theory of the list -- as a mapping;
- restructuring of Section E: the idea is to combine the Hansen reading of Popper; the aspects of the relevance model (logistic regression,  much data, a/b experiments, good features, substitution/complementarity), and the social theory of the list
- popper quotes in order:

> the introduction of propensities amounts to generalizing and extending the idea of forces again  14

> they should be regarded as _inherent in a situation_ 14

> we need _a calculus of relative or conditional probabilities_ as opposed _a calculus of absolute probabilities_ 16

> in many kinds of events this is not the case, and the propensities cannot be measured because the relevant situation changes and cannot be repeated 17

> in our real changing world, the situation and, with it, the possibilities, and thus the propensities, change all the time. They may certainly change if we, or any other organisms,  _prefer_ one possibility to another; or if we _discover_ a possibility where we have not seen one before 17

> experiments. ... by creating, at will, artificial conditions that either exclude, or reduce to zero, all the interfering and disturbing propensities 23

> What may happen in the future .. is, to some extent, open. There are many possibilities trying to realize themselves, but few of them have a very high propensity, given the initial conditions. 22

- an experiment for the conditional probability argument using logistic regression;
    - simulated 500k customers in terms of Taking Part data
    - simulate 52 weeks of purchase history using Tesco API data
    - build a separate model for each customer; 
    - see how that model can recommend substitutions or complementarities; these are 'new possibilities'   
- TODO: move the personalization discussion higher up, perhaps in the introduction
- rough reconstruction is ~4.2k, but haven't brought in the experiments much

### 2017-07-18 reconstruction.rmd
restructuring last section

This paper is a philosophical reconstruction [@Dewey_2017] of an online grocery ordering system. The reconstruction is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how recommendations were created for online grocery orders. Given this highly mediated and limited encounter with a recommender system and one of its designers, the paper constructs a multi-layered account. It draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping. It conducts several code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some important elements of the system in question.  It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] in order to identify and map the functional statements and practices underpinning the knowledge and order generated by such systems over time. Not least, I draw on the philosophy of science and media theory to frame an engagement with data, algorithms and recommendations around probabilities and their enactments. Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Rabinow_2004; @Bogost_2012], through this combination of approaches,  the paper itself provisionally allows its own reconstruction to be shaped by an experiential encounter with the object it describes. At times, this mode of empirical philosophical engagement with its objects may well seem overly concerned with the technical minutiae of the grocery ordering system, and lacking in  sociological rigour or empirical evidence. With some forbearance on the part of readers, however, the main philosophical argument of the paper should resonate sociologically. It concerns social order, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities play a more pervasive role.

The main argument of the paper concerns how  probability operates in Big Data situations. While calculations of probability have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on a much thicker situational texture. Operational probabilities differ from epistemic probabilities in their position in the world, as part of infrastructures, transactions, and event-structures. Taking their complex siting into account, and accepting their _physical_ reality, we might come to understand the world they belong to as what the philosopher of science Karl Popper understood as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. Given this understanding of operational probability, our understanding of what is enacted in contemporary data analytics, especially in association with platforms [@Gillespie_2010] might shift.

The broad implication of the argument concerns how we think about social order or regularities: proponents and critics of Big Data present the advent of large volumes of data and analytical techniques (machine learning in particular) as transforming organisations, institutions, economic processes and social life. I suggest too that a transformation or conversion event is occurring, but one which might open up the possibility of experiencing the world differently, and conceiving of social ordering processes somewhat afresh.  Even if the argument is centred on an eminently social practice such as shopping, the analysis of how a world of propensities continuously and unevenly converts into realizations applies across the board, from the most physical to the most ideal or abstract settings. The average everydayness of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- particle physics for instance -- do not.

@@

The shelves of large contemporary supermarket are the endpoint of global logistic supply chains, in all their logistical, value-transforming and brand-mediated hypercomplexity (see [@Neilson_2012; @Tsing_2009]).  Groceries imply a planetary geography of agriculture, industry, transport, communication and financialisation animated by flows of labour and capital. Encounters between this hyper-complex commodity-geography and people, even in the familiar confines of a supermarket, are no simple matter, either for shoppers or for supermarket operators such as Tesco.

@@

The conversion of  collaborative filtering into predictive personalisation appeared in Patel's presentations. Patel describes the move from 'a rules-based system'  to  a 'personal relevance model'. Each term in that phrase -- personal, relevance and model -- has some weight in the conversion event. We can assume that what Patel describes as  the 'rule-based system' refers  to the `apriori` algorithm, developed  by Rakesh Agrawal and Ramakrishnan Srikanti at IBM Research Alameda in the early 1990s [@Agrawal_1994]. A now-classic approach to 'market basket analysis,' it was listed as one of the top ten data mining algorithms in a survey conducted amongst data miners [@Wu_2008] and usually attracts a chapter in data-mining and machine learning textbooks (e.g. [@Hastie_2009]).

@@

@@

>>>>

This paper is a philosophical reconstruction [@Dewey_2017] of an online grocery ordering system undergoing an upgrade. The reconstruction is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how recommendations were created for online grocery orders. Starting from this highly mediated and fragmentary encounter with a recommender system and one of its designers, the paper explores the problem the system addresses.  It draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping and shopping lists. It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] in order to identify and map important functional statements and practices configuring the knowledge and order generated by such systems over time. It conducts several code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some prototypical elements of the system in question.  In doing this, it draws  on the philosophy of science and media theory to frame an engagement with data, algorithms and recommendations around probabilities and their worldly enactments.

Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Marcus_2014; @Bogost_2012], through this combination of approaches,  the paper allows its own reconstruction to be provisionally shaped by an experiential encounter with the object it describes. At times, this mode of empirical philosophical engagement with its objects may well seem overly concerned with the technical minutiae of the grocery ordering system, and lacking in  sociological rigour or empirical evidence. With some forbearance on the part of readers, however, the main philosophical argument of the paper should resonate sociologically. It concerns social order, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities play a more pervasive role.

The main argument of the paper concerns how  probability operates in Big Data situations. While calculations of probability have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on a much thicker situational texture. In recent research and opinion on  data practice, personalization has been seen as the linchpin of predictive modelling. The reconstruction of the recommender system suggests a different way of seeing these transformations.   Operational probabilities differ from epistemic probabilities in their position in the world, as part of infrastructures, transactions, and event-structures. Taking their complex siting into account, and accepting their _physical_ reality, we might come to understand the world operative probabilities belong to as what the philosopher of science Karl Popper understood as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. Given this understanding of operational probability, our understanding of what is enacted in contemporary data analytics, especially in association with platforms [@Gillespie_2010] where the operationality of probabilities are heightened,  might shift.

The broad implication of the argument concerns how we think about social order or regularities amidst change or transformation. Proponents and critics of Big Data present the advent of large volumes of data and analytical techniques (machine learning in particular) as transforming for better or worse organisations, institutions, economic processes and social life. Personalization has been a pivotal element, they argue, of this transformation.  I suggest too that a transformation or conversion event is occurring, but one which might also, even in the midst of much ongoing personalization,  open up the possibility of experiencing the world differently, and conceiving of social ordering processes somewhat afresh. Even if the argument is centred on shopping, the analysis of how a world of propensities continuously and unevenly converts into realizations applies across the board, from the most physical to the most ideal or abstract settings. The average everydayness of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- particle physics for instance -- do not.

@@

[TBA: THIS PARA could be moved down] The shelves of large contemporary supermarket are the endpoint of global logistic supply chains, in all their logistical, value-transforming and brand-mediated hypercomplexity (see [@Neilson_2012; @Tsing_2009]).  Groceries imply a planetary geography of agriculture, industry, transport, communication and financialisation animated by flows of labour and capital. Encounters between this hyper-complex commodity-geography and people, even in the familiar confines of a supermarket, are no simple matter, either for shoppers or for supermarket operators such as Tesco.

@@

[BREAK HERE]

@@

## The realization of propensities

>the introduction of propensities amounts to generalizing and extending the idea of forces again  14

>they should be regarded as _inherent in a situation_ 14

The conversion of  collaborative filtering into predictive personalisation appeared throughout  Patel's presentations. Patel described a move or 'conversion'  from 'a rules-based system'  to  a 'personal relevance model' as the basis of the recommender system. Each term in that phrase -- personal, relevance and model -- has some weight in the conversion event. Again, dealing with the sometimes telegraphic style of machine learning presentations means making some assumptions. It is reasonable to assume that what Patel describes as  the 'rule-based system' refers  to the extremely well-known `apriori` algorithm, developed  by computer scientists Rakesh Agrawal and Ramakrishnan Srikanti working at IBM Research Alameda in the early 1990s [@Agrawal_1994]. A now-classic approach to 'market basket analysis,' it was listed as one of the top ten data mining algorithms in a survey conducted amongst data miners [@Wu_2008] and usually attracts a chapter in data-mining and machine learning textbooks (e.g. [@Hastie_2009]). The interest of `apriori` for our purposes is that it relies on massive datasets, but does not personalize its recommendations.

The media theorist Mark Hansen has recently suggested that 'predictive analytics are discoveries of micrological propensities that are not directly correlated with human understanding and affectivity and that do not by themselves cohere into clearly identifiable events' [@Hansen_2015a, 111-112]. Hansen's account differs from many widely shared views of big data. In the face of the so much personalization (and Patel's presentation exemplifies this), Hansen suggests a somewhat processes of impersonalization.  If  much predictive practice attempts to elicit 'micrological propensities' from data,   Hansen, drawing on the work of the philosopher of science Karl Popper,  links data to things via probabilities:

>Whatever explanatory and causal value predictive analytics of large datasets have is, I suggest, ultimately rooted in this ontological transformation whereby probabilities are understood to be expressions of the actual propensity of things [@Hansen_2015a, 120]

Hansen theorises big data or predictive analytics as an 'ontological transformation' that deploys probabilities in an evermore closely woven and encompassing expression of animated, eventful, propensities of things. Predictive analytics links probabilities as calculations to propensities or the mutable associative agencies of things. The practical and indeed empirical  question is whether such transformations or 'conversion' in relations between probabilities and 'the propensities of things'  can be detected and articulated in the prosaic setting of shopping lists and recommender systems.[^20.1] The series of prototyping reconstructions I am about to discuss will explore the tenability of this rather radical alternative to regular narratives of personalization.

[^20.1]: I will argue that 'conversion' is a preferable term for these changes.  It highlights  re-orientations in subjects, experience, things, numbers and structures that ranges more widely than Hansen's 'ontological transformation' formulation. Not much hinges on the choice of terms, but 'conversion' happens to be the term used by Patel.

## Apriori as conditional probabilities

> we need _a calculus of relative or conditional probabilities_ as opposed _a calculus of absolute probabilities_ 16

```{r arules-ex, cache=TRUE, fig.cap='Frequency of association betwene items in the `Grocery` dataset', message=FALSE, echo=FALSE, fig.margin=TRUE}
library(knitr)
library(arules)
library(arulesViz)
library(datasets)
data(Groceries)
itemFrequencyPlot(Groceries,topN=10,col=rainbow(10), ntype="absolute",  cex.axis = 0.55)
```

```{r arules-ex2}
options(digits=2)
rules<- apriori(data=Groceries, parameter=list(supp=0.001,conf = 0.15,minlen=2), appearance = list(default="rhs",lhs="whole milk"), control = list(verbose=FALSE))
rules_sorted <- sort(rules, decreasing=TRUE,by="confidence")
kable(inspect(rules_sorted[1:5]), caption='The first five association rules for the `Groceries` dataset')
```

The `apriori` algorithm finds sets of items that commonly occur together in transactions.  These commonly occurring sets are expressed as 'association rules.' For instance, using the `Groceries` dataset in the `R` package `arules,` we can readily see how Walmart *might* have discovered   the fabled association between diapers and beer.[^149.1] Figure \@ref(fig:arules-ex) shows how often common items appear. `Whole milk` appears most frequently.  The association rule-based model for the `Groceries` dataset illustrates some characteristic ordering practices of 1990s-style data-mining approaches to lists (the `Groceries` dataset  was acquired from a 'local German supermarket' [@Hahsler_2006a]). The association between items is expressed in the form of _rules_ not a model. These rules, a term derived from older decision support literature, define sets of associations between items. Table \@ref(tab:arules-ex2) shows that the first rule has a stronger association than the second. Milk and vegetables is more likely an association than milk and buns.  A rules-based recommender system should, therefore, might put them in that order in the 'Have You Forgotten' list.

[^149.1]: One of the founding narratives of big data customer analytics, the beer-diaper association was not, according to [@Power_2002] found through data mining at all.

Although association rules and the `apriori` algorithm present social order in terms of sets of frequency-weighted associations, an `apriori`-based recommender system contains no statistical model. It also no grasp on personal relevance since the associations it defines are only between things. Put in terms of the big data conversion, the rules-based system is not predictive, or personal. The `apriori` algorithm might be seen as an attempt to grasp the excess of things in a Tesco supermarket by identifying subsets whose closure derives from the associations between things that people construct (through buying habits, by virtue of a shopping list, etc.). In association rules, personhood or experience is irrelevant. The openness of these rules -- the `apriori` algorithm can product any number of rules of diminishing strength -- points towards the propensity of things to find themselves together in shopping baskets, but cannot cannot individuate that propensity beyond demographic quantiles and categories.

## `Apriori` meets the API

Even if `apriori`  elicits associations between things, it struggles with the propensity of commodities to multiply. A simple illustration of the combinatorial problem faced by recommender system can be developed using the `Grocery` dataset.  If we take all the items in the `Grocery` dataset and paste them into the 'shopping list' box on the Tesco grocery website (or as I side, run them as searches on the TescoLab Product API), each of the 169 items in the dataset yields dozens and sometimes thousands of products from Tesco online.

```{r tesco-ex, echo=FALSE, cache=TRUE, fig.cap ='Tesco surplus',  fig.margin=TRUE}
    library(ggplot2)
    tesco = read.csv('data/reference_data/tesco_groceries.csv')
    total = sum(tesco$actual, na.rm=TRUE) 
    tesco = na.omit(tesco)
    tesco$actual = as.integer(tesco$actual)
    ggplot(tesco[tesco$actual>50,], aes(x=labels, y=actual, fill=labels)) +geom_bar(stat='identity') + coord_flip() + ggtitle('Tesco grocery items with more than 50 products') + theme(legend.position="none", axis.text=element_text(size=7))
```

Items in the `Groceries` dataset proliferate into a Tesco's list of branded variations. The 169 items of the `Groceries` dataset  expand into roughly `r total` Tesco items (see Figure \@ref(fig:tesco-ex)).  There is, I would suggest, a logistic propensity in things that recommender systems encounter in the ways of global supply chains. The multiple listed  in the `Grocery` dataset becomes more open in this setting through new impersonal intersections that must be bounded and ordered. The proliferation of things on the shelves of supermarket or grocery warehouse confronts produces a combinatorial problem for data miners. `r total` products (actually Patel mentioned 200,000 products) can be combined in a vast number of ways. If a typical shopping list has 20 items, then there are `r format(choose(total, 20), 2)` possible lists. Some of this vast number of possible shopping lists are highly unlikely in terms of typical purchasing patterns, but potentially valuable as recommendations.

##  The list in its mapping to probabilities

Conditional probability; the models
[TBA - section on the role of probability; and logr]

## Repeating sufficiently often to measure

52 weeks of data
> in many kinds of events this is not the case, and the propensities cannot be measured because the relevant situation changes and cannot be repeated 17

##  Platform Experiments that reduce interfering propensities

> experiments. ... by creating, at will, artificial conditions that either exclude, or reduce to zero, all the interfering and disturbing propensities 23

## The openness of the data -- new possibilities

> What may happen in the future .. is, to some extent, open. There are many possibilities trying to realize themselves, but few of them have a very high propensity, given the initial conditions. 22

@@

In their mundane intersectional practices of ordering, recommender systems show how  predictive practices affect existing forms of social order in the world. Rather than directly positing an ontological transformation (as Hansen does), or pointing to the erosion of social order (as Turow does), I have explored some of the conversion events associated with a shift from a data-mining rule-based system to a predictive recommender system. The DunnHumby model and the Tesco 'Have You Forgotten?' recommendations present many of the traits of the conversion processes associated with big data. Rather than discontinuous changes, the conversion of the recommender system via a relevance model suggests an ongoing problematization of propensities. We have seen several types of conversion events: ongoing transformation of models and infrastructures, the experimentalisation  of recommendations, and the broader diffusion of a calculus of probabilities into a fibrous weave of propensities and conditionalities, warping over time. The predictive assemblages connect logistical ensembles to the negotiation, compromises, rituals and local orderings of everyday life.

I have been suggesting that predictive listing in the form of open-ended recommendation changes social order. It establishes a common coding between machinic and ritual repetition (as exemplified in the shopping list).  It also suggests some ways in which repetition or reiteration  can also participate in changing and emergent order. By understanding the world in terms of propensities, we have a chance  to engage with contemporary predictive processes in ways that both acknowledge their relentless and exhausting pressure to consume more and retain some chance of inhabiting the world differently.

Rather than understanding prediction as the closed form of an ontologically open world ('the propensities of things'), the DunnHumby relevance model, in all its statistical, infrastructural, organisational and banal complexity, suggests that the predictive conversion remains constitutively incomplete. Shopping lists in all their banal yet primary orderings are fertile terrain for prediction because they are incomplete, somewhat open and mutable forms.  The constitutive incompleteness and openness of lists are very closely entangled with ongoing processes of ordering and othering, and at the same time, can accommodate predictive operations.   Lists undergoing a 'conversion' experience as they become more probabilistic, but their heterogeneous character as operations caught up in models, infrastructures, experiments, and the flow of commodities means that they can never form a closed set or dataset. The problem of 'constructing good features,' a problem that any predictive model must address, persists as the site of experiments in partly open structures.

A constitutive openness to new associations, new combinations, new groupings or actors is integral to the conversion events associated with contemporary data practice. In their account of the Netflix recommender system, Ted Hallinan and Peter Striphas make the following argument:

> These systems, and more importantly their algorithms, play a critical role in deciding which articles (or parts thereof) gain admission to the cultural realm, and in what form. Their doing so thus points in the direction of another universe of reference—a court of algorithmic appeal in which objects, ideas, and practices are heard, cross-examined, and judged independently, in part, of human beings [@Hallinan_2014, 13]

Despite the somewhat troubling reference to 'another universe,' as if Netflix or DunnHumby's recommender systems occupy a different reality, Hallinan and Striphas usefully and aptly highlight how such systems  mingle  objects, ideas and practices.

Something similar happens to shopping in the Dunnhumby relevance model. In the customer-product association matrix, in the product-product similarity matrix, in the 14,000 variables included in the logistic regression model, a 'universe of reference' takes shape. The shopping list, the commonplace inscription of normative and actual social order, takes on a complex temporal-material structure that mixes prediction, infrastructural constraints, the individual characteristics of list-makers, statistical validations, database architectures and above all the surplus sensibility associated with  circulation and consumption of commodities. Shopping lists undergoing a conversion experience as they become more probabilistic, but the idea that probabilities are the closed form of propensities does not account for the distribution of probabilities in relations between things, between people, between people and things, and between things and people.
